### 2025-08-23 sports-cognitive-psychology.qmd
20250823 - copyediting

The findings described above might suggest that more attention leads to better performance.
Based on that assumption, we should tell a Quarterback who has thrown multiple interceptions during a game to pay more attention to his throwing mechanics when throwing the ball, right?
However, it is not always the case that devoting more attention to a task leads to better performance.
There are two primary competing theories for "choking" in pressure situations in sports.
"Choking" refers to underperforming relative to one's ability [@Beilock2001].
Distraction theories posit that the pressure situation creates a distracting environment that shifts one's attention to cues that are irrelevant to the task, such as one's worries about the situation and its consequences [@Beilock2001].
An alternative possibility, based on explicit monitoring theory, is that pressure situations raise one's self-awareness and anxiety about performing well, which increases one's attention to the step-by-step procedures involved in skill processes, which may disrupt well-learned or proceduralized performance [@Beilock2001].
@Beilock2001 found that, among automatic tasks (e.g. putting in golf among highly trained individuals), focusing too much on the task can actually hinder performance, consistent with explicit monitoring theory.
When a task is highly practiced or automatic, excessive attention can hurt performance.
That is, more attention can lead to worse performance.
This is called "choking under pressure" and is based on explicit monitoring theory.

Explicit monitoring theories posit that pressure situations raises one's self-awareness and anxiety about performing well, which increases one's attention to the stop-by-step procedures involved in skill processes [@Beilock2001].
For highly practiced experts, automatic skill tasks might include tasks like hitting/putting a golf ball, pitching a baseball, throwing/kicking a football, shooting free throws in basketball, shooting arrows in archery, etc.
Sports players often refer to highly practiced tasks as involving "muscle memory."

>>>>

The findings described above might suggest that more attention leads to better performance.
Based on that assumption, we should tell a Quarterback who has thrown multiple interceptions during a game to pay more attention to his throwing mechanics when throwing the ball, right?
However, it is not always the case that devoting more attention to a task leads to better performance.
There are two primary competing theories for "choking" in pressure situations in sports.
"Choking" refers to underperforming relative to one's ability [@Beilock2001].

Distraction theories posit that the pressure situation creates a distracting environment that shifts one's attention to cues that are irrelevant to the task, such as one's worries about the situation and its consequences [@Beilock2001].

An alternative possibility, based on explicit monitoring theory, is that pressure situations raise one's self-awareness and anxiety about performing well, which increases one's attention to the step-by-step procedures involved in skill processes, which may disrupt well-learned or proceduralized performance [@Beilock2001].
@Beilock2001 found that, among automatic tasks (e.g. putting in golf among highly trained individuals), focusing too much on the task can actually hinder performance, consistent with explicit monitoring theory.
For highly practiced experts, automatic skill tasks might include tasks like hitting/putting a golf ball, pitching a baseball, throwing/kicking a football, shooting free throws in basketball, shooting arrows in archery, serving in tennis, etc.
Sports players often refer to highly practiced tasks as involving "muscle memory."
When a task is highly practiced or automatic, excessive attention can hurt performance.
That is, more attention can lead to worse performance.
This is called "choking under pressure" and is based on explicit monitoring theory.
